Why Real-World Subgraph Isomorphism is Easy, \\ and What This Means for Graph Databases

### 2016-10-20 paper/paper.tex
James
Feedback from Christine

\author{\name Ciaran McCreesh \thanks{This work was supported by the Engineering and Physical Sciences
            Research Council [grant number EP/K503058/1]} \email c.mccreesh.1@research.gla.ac.uk \\
       \name Patrick Prosser \email patrick.prosser@glasgow.ac.uk \\
       \name James Trimble \\
   \addr University of Glasgow, Glasgow, Scotland}
\maketitle

\begin{abstract}
    Despite subgraph isomorphism being \NP-complete, modern algorithms can operate comfortably on
    real-world graphs with thousands of vertices. However, they cannot handle arbitrary instances of
    this size. We show how to generate ``really hard'' random instances for subgraph isomorphism
    problems, which are computationally challenging with only a couple of hundred vertices. For the
    non-induced version of the problem, these instances lie on a satisfiable / unsatisfiable phase
    transition, whose location we can predict; for the induced variant, much richer behaviour is
    observed, and constrainedness gives a better measure of difficulty than does proximity to a
    phase transition. In both cases, we can use this to understand and improve variable and value
    ordering heuristics.

    When labels are introduced, as commonly seen in graph databases, additional rich behaviour
    arises.  We show that when only slightly better algorithms are used, all commonly used benchmark
    suites for evaluating graph database systems become trivial, and that existing notions of
    difficulty are based almost entirely upon unnecessarily poor performance from overly simplistic
    subgraph isomorphism algorithms. With this in mind, we argue that the widely researched ``filter
    / verify'' indexing technique used in graph databases is a flawed design based upon a repeated
    misunderstanding of what makes subgraph isomorphism hard. We explain why this technique would
    not be beneficial, even with harder instances than those currently used, when paired with any
    reasonable modern subgraph isomorphism algorithm.
\end{abstract}

@@

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from a given
pattern graph to a given target graph which preserves adjacency---in essence, we are ``finding a
copy of'' the pattern inside the target. The \emph{induced} variant of the problem additionally
requires that the mapping preserve non-adjacency, so there are no ``extra edges'' in the copy of the
pattern that we find. We illustrate both variants in \cref{figure:sip}.  Although these problems
are \NP-complete, modern practical subgraph isomorphism algorithms can handle problem instances
with many hundreds of vertices in the pattern graph, and up to ten thousand vertices in the target
graph \cite{Cordella:2004,Solnon:2010,Audemard:2014,McCreesh:2015}, leading to successful
application in areas such as computer vision \cite{Damiand:2011,Solnon:2015}, biochemistry
\cite{Giugno:2013,Carletti:2015}, and pattern recognition \citep{Conte:2004}.  In particular,
a labelled version of subgraph isomorphism is one of the key components in supporting complex
queries in graph databases---typically, these systems store a fixed set of target graphs, and for a
sequence of pattern queries, they must return every target graph which contains that pattern.  One
common approach to this problem is to combine a subgraph isomorphism algorithm with an indexing
system in a so-called ``filter / verify'' model, where invariants are used to attempt to pre-exclude
unsatisfiable instances to avoid the cost of a subgraph isomorphism call. In this context, the terms
\emph{matching} and \emph{verification} are often used for the subgraph isomorphism step (or a
slightly broader problem, for example permitting wildcards).

@@

Despite these successes, subgraph isomorphism algorithms cannot handle \emph{arbitrary} instances
involving hundreds or thousands of vertices. Experimental evaluations of subgraph isomorphism
algorithms are usually performed using a mix of real-world graphs, graphs that encode biochemistry
and computer vision problems, and randomly generated graph pairs. Using random instances to evaluate
algorithm behaviour can be beneficial, because it provides a way of generating many instances
cheaply, and reduces the risk of over-fitting when tuning design parameters. The random instances
used in each case came from common datasets \citep{DeSanto:2003,Zampelli:2010}, which were generated
by taking a random subgraph of a random (Erd\H{o}s-R\'enyi, scale-free, bounded degree, or mesh)
graph and permuting the vertices. Such instances are guaranteed to be
satisfiable---\citet{Anton:2009} exploited this property to create large sets of random satisfiable
boolean satisfiability instances.  This has been by far the most common approach used so far to
generate random subgraph isomorphism instances so far, meaning existing benchmark suites contain
relatively few non-trivial unsatisfiable instances.  Also, the satisfiable instances tend to be
computationally fairly easy, with most of the difficulty being in dealing with the size of the
model. This has lead to bias in algorithm design, to the extent that some proposed techniques will
\emph{only} work on satisfiable instances \citep{Battiti:2007}.

@@

@@

When labels on vertices are introduced, as is commonly seen in graph database systems, richer
behaviour emerges. We look at some of the datasets commonly used to test graph database systems,
which the literature suggests are extremely hard to solve. Our theory predicts that these instances
should be trivial, and indeed we see that the entire perceived difficulty of every dataset we
investigate comes down to the widespread use of a particular subgraph isomorphsim algorithm which
does not use a theoretically-strong variable ordering heuristic. Our second contribution is to show
that the use of the right variable ordering heuristic, even without the more sophisticated recent
advances in subgraph algorithms (which have been largely neglected outside of the artificial
intelligence community), makes the entire graph database filter / verify paradigm unnecessary.
Finally, we explain why filter / verify cannot be beneficial even on more challenging instances
except if an extremely poor choice of subgraph isomorphism algorithm is made.

@@

@@

A \emph{non-induced subgraph isomorphism} from a graph $P$ (called the \emph{pattern}) to a graph
$T$ (called the \emph{target}) is an injective mapping $i$ from $\operatorname{V}(P)$ to
$\operatorname{V}(T)$ which preserves adjacency---that is, for every adjacent $v$ and $w$ in
$\operatorname{V}(P)$, the vertices $i(v)$ and $i(w)$ are adjacent in $T$. If vertices are labelled,
the mapping must preserve labels, so $\ell(v) = \ell(i(v))$. An \emph{induced subgraph isomorphism}
additionally preserves non-adjacency---that is, if $v$ and $w$ are not adjacent in $P$, then $i(v)$
and $i(w)$ must not be adjacent in $T$. We use the notation $i : P \rightarrowtail T$ for a
non-induced isomorphism, and $i : P \hookrightarrow T$ for an induced isomorphism. Observe that an
induced isomorphism $i : P \hookrightarrow T$ is a non-induced isomorphism $i : P \rightarrowtail T$
which is also a non-induced isomorphism $\shortoverline{i} : \shortoverline{P} \rightarrowtail
\shortoverline{T}$.

@@

    \caption{Behaviour of algorithms on the non-induced variant. For each plot, the x-axis is the
        pattern edge probability and the y-axis is the target edge probability, both from 0 to 1.
        Along the top row, we show the proportion of instances which are satisfiable; the white
        bands shows the phase transitions, and the black lines are our predictions of where the
        phase transition will occur. On the final three rows, we show the number of search nodes used by the
        Glasgow, LAD and VF2 solvers; the dark regions indicate ``really hard'' instances.}
    \label{figure:non-induced}
\end{figure}

@@

Suppose our labels are drawn randomly from a set $L = \{ 1, \ldots, k \}$, where $k$ is reasonably
small compared to $p$. We can partition the pattern vertices by label into disjoint sets \[\{
    \operatorname{V}(P)\vert_1 = \{ v \in \operatorname{V}(P) : \ell(v) = 1 \}, \ldots,
    \operatorname{V}(P)\vert_k \}\textnormal{\,,}\] each of which is expected to contain
$\nicefrac{p}{k}$ vertices. Similarly, we may partition the target vertices into disjoint sets $\{
    \operatorname{V}(T)\vert_1, \ldots, \operatorname{V}(T)\vert_k \}$.

Without labels, there are $t^{\underline{p}} = t \cdot (t - 1) \cdot \ldots \cdot (t - p + 1)$
possible injective assignments of target vertices to pattern vertices.  With labels, observe that
for any label $x$, vertices in $\operatorname{V}(P)|_x$ may only be mapped to vertices in
$\operatorname{V}(T)|_x$.  Thus for each label $x$, we have an expected $\nicefrac{p}{k}$ variables,
each of whose domains contain $\nicefrac{t}{k}$ values. We would like to say that the size of the
state space is now \[ |S| = \left((\nicefrac{t}{k})^{\underline{\nicefrac{p}{k}}}\right)^{k}
\textnormal{,} \] but to do this we must state what $a^{\underline{b}}$ means when $b$ is fractional.
A reasonable continuous extension may be obtained by taking \[ |S| = \left(\frac{\Gamma\left(\nicefrac{t}{k}
+ 1\right)}{\Gamma\left(\nicefrac{t}{k} - \nicefrac{p}{k} + 1\right)}\right)^{k} \textnormal{\,.} \]

@@

So how good are these predictions? We draw some heatmaps in \cref{figure:labels}. For small numbers
of labels, our predictions are slightly better than in the unlabelled case: there seems to be less
of a variance problem for very dense patterns. Even as the numbers of labels becomes relatively
large, the prediction of the phase transition still occurs in the right place, but we start to see
sporadic unsatisfiable instances deep inside the satisfiable reason. With sufficiently many labels,
we get a kind of phase transition from ``all unsatisfiable'' to ``mixed satisfiable and
unsatisfiable''.

@@

Without using the equation, \citet{DBLP:conf/sigmod/ChengKNL07} state that since subgraph
isomorphism is NP-complete, processing by a sequential scan is infeasible. They introduce new
filtering techniques to try to avoid the subgraph isomorphism step. A similar claim is made by
\citet{DBLP:conf/icde/ZhangHY07} in a introduction of another indexing technique: ``obviously it is
inefficient to perform a sequential scan on every graph in the database, because the subgraph
isomorphism test is expensive''.

>>>>

\author{\name Ciaran McCreesh \thanks{This work was supported by the Engineering and Physical Sciences
           Research Council [grant number EP/K503058/1]} \email c.mccreesh.1@research.gla.ac.uk \\
       \name Patrick Prosser \email patrick.prosser@glasgow.ac.uk \\
       \name James Trimble \thanks{This work was supported by the Engineering and Physical Sciences
           Research Council [grant number EP/M508056/1]} \email j.trimble.1@research.gla.ac.uk \\
   \addr University of Glasgow, Glasgow, Scotland}
\maketitle

\begin{abstract}
    Despite subgraph isomorphism being \NP-complete, modern algorithms can operate comfortably on
    most real-world graphs with thousands of vertices. However, they cannot handle arbitrary instances of
    this size. We show how to generate ``really hard'' random instances for subgraph isomorphism
    problems, which are computationally challenging with only a couple of hundred vertices. For the
    non-induced version of the problem, these instances lie on a satisfiable / unsatisfiable phase
    transition, whose location we can predict; for the induced variant, much richer behaviour is
    observed, and constrainedness gives a better measure of difficulty than does proximity to a
    phase transition. In both cases, we can use this to understand and improve variable and value
    ordering heuristics.

    When labels are introduced, as commonly seen in real-world applications and in graph databases,
    additional rich behaviour arises.  We show that when only slightly better algorithms are used,
    all commonly used benchmark suites for evaluating graph database systems become trivial, and
    that existing notions of difficulty are based almost entirely upon unnecessarily poor
    performance from overly simplistic subgraph isomorphism algorithms. With this in mind, we argue
    that the widely researched ``filter / verify'' indexing technique used in graph databases is a
    flawed design based upon a repeated misunderstanding of what makes subgraph isomorphism hard. We
    explain why this technique would not be beneficial, even with harder instances than those
    currently used, when paired with any reasonable modern subgraph isomorphism algorithm.
\end{abstract}

@@

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from the
vertices of a given pattern graph to the vertices of a given target graph which preserves
adjacency---in essence, we are ``finding a copy of'' the pattern inside the target. The
\emph{induced} variant of the problem additionally requires that the mapping preserve non-adjacency,
so there are no ``extra edges'' in the copy of the pattern that we find. We illustrate both variants
in \cref{figure:sip}.  Although these problems are \NP-complete, modern practical subgraph
isomorphism algorithms can handle problem instances with many hundreds of vertices in the pattern
graph, and up to ten thousand vertices in the target graph
\cite{Cordella:2004,Solnon:2010,Audemard:2014,McCreesh:2015}, leading to successful applications in
areas such as computer vision \cite{Damiand:2011,Solnon:2015}, biochemistry
\cite{Giugno:2013,Carletti:2015}, and pattern recognition \citep{Conte:2004}.  In particular, a
labelled version of subgraph isomorphism is one of the key components in supporting complex queries
in graph databases---typically, these systems store a fixed set of target graphs, and for a sequence
of pattern queries, they must return every target graph which contains that pattern.  One common
approach to this problem is to combine a subgraph isomorphism algorithm with an indexing system in a
so-called ``filter / verify'' model, where invariants are used to attempt to pre-exclude
unsatisfiable instances to avoid the cost of a subgraph isomorphism call. In this context, the terms
\emph{matching} and \emph{verification} are often used for the subgraph isomorphism step (or a
slightly broader problem, for example permitting wildcards).

@@

Despite these successes, subgraph isomorphism algorithms cannot handle \emph{arbitrary} instances
involving hundreds or thousands of vertices. Experimental evaluations of subgraph isomorphism
algorithms are usually performed using a mix of real-world graphs, graphs that encode biochemistry
and computer vision problems, and randomly generated graph pairs. Using random instances to evaluate
algorithm behaviour can be beneficial, because it provides a way of generating many instances
cheaply, and reduces the risk of over-fitting when tuning design parameters. The random instances
used in each case came from common datasets \citep{DeSanto:2003,Zampelli:2010}, which were generated
by taking a random subgraph of a random (Erd\H{o}s-R\'enyi, scale-free, bounded degree, or mesh)
graph and permuting the vertices. Such instances are guaranteed to be
satisfiable---\citet{Anton:2009} exploited this property to create large sets of random satisfiable
boolean satisfiability instances.  This has been by far the most common approach used so far to
generate random subgraph isomorphism instances so far, meaning existing benchmark suites contain
relatively few non-trivial unsatisfiable instances (although some of the patterns in the instances
by \citeauthor{Zampelli:2010} have had extra edges added, making them unsatisfiable). Also, the
satisfiable instances tend to be computationally fairly easy, with most of the difficulty being in
dealing with the size of the model. This has lead to bias in algorithm design, to the extent that
some proposed techniques will \emph{only} work on satisfiable instances \citep{Battiti:2007}.

@@

?? This needs to be reworded if we include non-ER random graphs.

@@

When labels on vertices are introduced, as is commonly seen in real-world applications and in graph
database systems, richer behaviour emerges. We look at some of the datasets commonly used to test
graph database systems, which the literature suggests are extremely hard to solve. Our theory
predicts that these instances should be trivial, and indeed we see that the entire perceived
difficulty of every dataset we investigate comes down to the widespread use of a particular subgraph
isomorphsim algorithm which does not use a theoretically-strong variable ordering heuristic. Our
second contribution is to show that the use of the right variable ordering heuristic, even without
the more sophisticated recent advances in subgraph algorithms, makes the entire graph database
filter / verify paradigm unnecessary.  Finally, we explain why filter / verify cannot be beneficial
even on more challenging instances except if an extremely poor choice of subgraph isomorphism
algorithm is made.

@@

?? Check whether any of the real-world graphs we look at have loops.

@@

A \emph{non-induced subgraph isomorphism} from a graph $P$ (called the \emph{pattern}) to a graph
$T$ (called the \emph{target}) is an injective mapping $i$ from $\operatorname{V}(P)$ to
$\operatorname{V}(T)$ which preserves adjacency---that is, for every adjacent vertices $v$ and $w$
in $\operatorname{V}(P)$, the vertices $i(v)$ and $i(w)$ are adjacent in $T$. If vertices are
labelled, the mapping must preserve labels, so $\ell(v) = \ell(i(v)) \forall v \in
\operatorname{V}(P)$. An \emph{induced subgraph isomorphism} additionally preserves
non-adjacency---that is, if $v$ and $w$ are not adjacent in $P$, then $i(v)$ and $i(w)$ must not be
adjacent in $T$. We use the notation $i : P \rightarrowtail T$ for a non-induced subgraph
isomorphism, and $i : P \hookrightarrow T$ for an induced subgraph isomorphism. Observe that an
induced subgraph isomorphism $i : P \hookrightarrow T$ is a non-induced subgraph isomorphism $i : P
\rightarrowtail T$ which is also a non-induced subgraph isomorphism $\shortoverline{i} :
\shortoverline{P} \rightarrowtail \shortoverline{T}$.

@@

    \caption{Behaviour of algorithms on the non-induced variant. For each plot, the x-axis is the
        pattern edge probability and the y-axis is the target edge probability, both from 0 to 1.
        Along the top row, we show the proportion of instances which are satisfiable; the white
        bands show the phase transitions, and the black lines are our predictions of where the
        phase transition will occur. On the final three rows, we show the number of search nodes used by the
        Glasgow, LAD and VF2 solvers; the dark regions indicate ``really hard'' instances.}
    \label{figure:non-induced}
\end{figure}

@@

Suppose our labels are drawn randomly from a set $L = \{ 1, \ldots, k \}$, where $k$ is reasonably
small compared to $p$. By defining \[ \operatorname{V}(P)\vert_n = \{ v \in \operatorname{V}(P) :
\ell(v) = n \} \] to be the set of vertices with label $k$, we can partition the pattern
vertices by label into disjoint sets \[\{ \operatorname{V}(P)\vert_1, \ldots,
\operatorname{V}(P)\vert_k \}\textnormal{\,,}\] each of which is expected to contain
$\nicefrac{p}{k}$ vertices. Similarly, we may partition the target vertices into disjoint
sets $\{ \operatorname{V}(T)\vert_1, \ldots, \operatorname{V}(T)\vert_k \}$.

Without labels, there are $t^{\underline{p}} = t \cdot (t - 1) \cdot \ldots \cdot (t - p + 1)$
possible injective assignments of target vertices to pattern vertices.  With labels, observe that
for any label $x$, vertices in $\operatorname{V}(P)|_x$ may only be mapped to vertices in
$\operatorname{V}(T)|_x$.  Thus for each label $x$, we have an expected $\nicefrac{p}{k}$ variables,
each of whose domains contain $\nicefrac{t}{k}$ values. We would like to say that the size of the
state space is now \[ |S| = \left((\nicefrac{t}{k})^{\underline{\nicefrac{p}{k}}}\right)^{k}
\textnormal{,} \] but to do this we must state what $a^{\underline{b}}$ means when $b$ is
fractional. The gamma function $\Gamma(n)$ is equal to $(n - 1)!$ for integers $n \ge 1$, but is also
defined for positive real numbers, obeying the identify $\Gamma(x + 1) = x\Gamma(x)$. By noting that
$t^{\underline{p}} = \nicefrac{t!}{(t - p)!}$, we may obtain a reasonable
continuous extension by taking \[ |S| = \left(\frac{\Gamma\left(\nicefrac{t}{k} +
1\right)}{\Gamma\left(\nicefrac{t}{k} - \nicefrac{p}{k} + 1\right)}\right)^{k} \textnormal{\,,} \]

@@

So how good are these predictions? We draw some heatmaps in \cref{figure:labels}. For small numbers
of labels, our predictions are slightly better than in the unlabelled case: there seems to be less
of a variance problem for very dense patterns. Even as the numbers of labels becomes relatively
large, the prediction of the phase transition still occurs in the right place, but we start to see
sporadic unsatisfiable instances deep inside the satisfiable region. With sufficiently many labels,
we get a kind of phase transition from ``all unsatisfiable'' to ``mixed satisfiable and
unsatisfiable''.

@@

Without using the equation, \citet{DBLP:conf/sigmod/ChengKNL07} state that since subgraph
isomorphism is NP-complete, processing by a sequential scan is infeasible. They introduce new
filtering techniques to try to avoid the subgraph isomorphism step. A similar claim is made by
\citet{DBLP:conf/icde/ZhangHY07} in an introduction of another indexing technique: ``obviously it is
inefficient to perform a sequential scan on every graph in the database, because the subgraph
isomorphism test is expensive''.
